Mergeable Types

### 2017-02-27 paper/mergeable-types.tex

\section{Mergeable Data Type}

@@

We now formally define what we mean by a mergeable type. The
definition serves as a guide to write merge functions for functional
data types, elevating them to the level of replicated data types.

A functional data type library is a tuple consisting of a type ($t$),
and a collection of functions ($f$, $g$, etc.) of type $t \rightarrow
t$ that are \emph{primitive morphisms}. A morphism ($P$, $Q$, etc.) is
either a primitive morphism, or an associative composition of
morphisms ($P \circ Q \circ R$). An object $B:t$ is called a
\emph{semantic successor} of $A:t$ (conversely, $A$ is called a
\emph{semantic ancestor} of $B$) if and only if there exists a
morphism $P$ such that $P(A) = B$.

@@

Rather than specifying merge ($M$) in terms of an LCA,
Definition~\ref{def:mergeable-type} specifies $M$ in terms of a pair
of morphisms.  This specification is useful to avoid the expression of certain
nonsensical merge functions, and thus enforce a useful branch-local
consistency property (Theorem~\ref{thm:branch-consistency}). For
instance, consider the counter data type from Sec.~\ref{sec:intro}. A
three-way merge function can be defined to return a constant
regardless of its arguments:
\begin{ocaml}
  let merge lca a b = 0
\end{ocaml}
Such a merge function allows the client to
witness the violation of the branch-local consistency property. For
example, the client may \C{sync} the counter, whose current local
value is $10$, and obtain $0$ as a result, thus witnessing a violation
of counter's monotonicity property. Fortunately,
Def.~\ref{def:mergeable-type} disallows this semantics. It is
impossible to define the function $M$ for the counter data type that
violates the monotonicity invariant, because no counter morphism
violates the invariant. In general, the specification of $M$
guarantees that the merge operation for a type $t$ preserves any
invariants (e.g., balancedness of a tree, sortedness of a list,
non-negativeness of an integer etc) that are preserved by all of $t$'s
morphisms.  The commutativity of the diagram in
Fig.~\ref{fig:pushouts} need not be enforced statically. It can
equally be verified at run-time by checking that the morphisms returned by $M$
map respective concurrent versions to the same value. The
operational semantics (Fig.~\ref{fig:opsem}) can be extended to this
effect.

@@

Merge functions for simple data types are easy to write. For instance,
the merge logic for counter is simple enough that it does not require
great effort to identify the $P$ and $Q$ that leda to the
concurrent counter versions, and transform them into $P'$ and $Q'$
such that the diagram in Fig.~\ref{fig:pushouts} commutes. For more
\begin{wrapfigure}{l}{.4\textwidth}
\centering
\includegraphics[scale=0.35]{Figures/pushouts-2}
\caption{Commutative diagram for Def.~\ref{def:mergeable-type-2}.}
\label{fig:pushouts-2}
\end{wrapfigure}
sophisticated data types, this process is more involved, and may require
the programmer to perform non-trivial reasoning. We now present an
alternative definition of a mergeable type that is stronger than
Def.~\ref{def:mergeable-type}, but also serves as a recipe to build
non-trivial merge functions.

\paragraph{Notation} As mentioned earlier, we distinguish between the
named functions defined by a data type library (primitive morphisms),
and their compositions (morphisms). A sequence (list) of primitive
morphisms is written $P^*$ or $Q^*$. The composition of a sequence of
primitive morphisms (written $\denot{P^*}$) is their right-to-left
composition, i.e., $\C{fold\_left}\;(\lambda f.\lambda g. f \circ
g)\; id\; P^*$, where $\C{fold\_left}$ has its usual OCaml type. The
length of a sequence $S$ is written $\length{S}$.

@@

The definition asserts the existence of two functions for a mergeable
type $t$. First, the function $W$, which computes the \emph{edit
distance} between an object $A:t$ and its semantic successor $B:t$.
The sequence of primitive morphisms it returns ($P^*$) is called an
\emph{edit script}. Replaying the edit script on $A$ produces $B$
(i.e., $\denot{P^*}(A) = B$). Second, the definition asserts the
existence of a function $M$, which performs an \emph{operational
transformation} of the primitive morphisms in an edit script. An
operational transformation of a primitive morphism $P$ w.r.t a
morphism $Q$ is another primitive morphism $P'$ such that for any
object $A:t$, $P'$ has the same \emph{effect} on $Q(A)$ as $P$ has on
$A$. We will now consider examples of mergeable types that reify the
notions of edit scripts and operational transformations.

>>>>

\subsection{Mergeable Data Type}

@@

We now formally define a mergeable type.  A functional data type
library is a tuple consisting of a type ($t$), and a collection of
functions ($f$, $g$, etc.) of type $t \rightarrow t$ that are
\emph{primitive morphisms}. A morphism ($P$, $Q$, etc.) is either a
primitive morphism, or an associative composition of morphisms ($P
\circ Q \circ R$). An object $B:t$ is called a \emph{semantic
successor} of $A:t$ (conversely, $A$ is called a \emph{semantic
ancestor} of $B$) if and only if there exists a morphism $P$ such that
$P(A) = B$.

@@

Rather than specifying merge ($M$) in terms of an LCA,
Definition~\ref{def:mergeable-type} specifies $M$ in terms of a pair
of morphisms.  This specification is useful to avoid the expression of certain
nonsensical merge functions, and thus enforce a useful branch-local
consistency property (Theorem~\ref{thm:branch-consistency}). For
instance, consider the counter data type from Sec.~\ref{sec:intro}. A
three-way merge function can be defined to return a constant
regardless of its arguments:
\begin{ocaml}
  let merge lca a b = 0
\end{ocaml}
Such a merge function allows the client to
witness the violation of the branch-local consistency property. For
example, the client may \C{sync} the counter, whose current local
value is $10$, and obtain $0$ as a result, thus witnessing a violation
of counter's monotonicity property. Fortunately,
Def.~\ref{def:mergeable-type} disallows this semantics. It is
impossible to define the function $M$ for the counter data type that
violates the monotonicity invariant, because no counter morphism
violates the invariant. In general, the specification of $M$
guarantees that the merge operation for a type $t$ preserves any
invariants (e.g., balancedness of a tree, sortedness of a list,
non-negativeness of an integer etc) that are preserved by all of $t$'s
morphisms.  The commutativity of the diagram in
Fig.~\ref{fig:pushouts} need not be enforced statically. It can
equally be verified at run-time by checking that the morphisms returned by $M$
map respective concurrent versions to the same value. The
\rulelabel{E-Pull-Wait} rule of the operational semantics
(Fig.~\ref{fig:opsem}) can be extended to this effect:
\begin{smathpar}
\begin{array}{c}
\RULE
{
  t\neq t' \spc
% \under{H}{v' \mbleto v} \spc
% \C{world}(H,t') \semsucceq \C{world}(H,t)\spc 
  v' \not\preceq v \spc
  (P',Q') = M(\C{lca}(H(t),H(t')), v, v') \spc
  Q'(v) = P'(v')
}
{
  (\pull)_t;H(t \mapsto (v,f)::m)(t' \mapsto (v',\_)::\_) ~\stepsto~
  (\pull)_t;H[t \mapsto (v_m,\C{MERGE}\; H(t'))::(v,f)::m]
}
\end{array}
\end{smathpar}

@@

Def.~\ref{def:mergeable-type} is immediately applicable to simple data
types like counter to explain why they are mergeable. The merge logic
for counter is simple enough that it does not require great effort to
identify the $P$ and $Q$ that lead to the concurrent counter versions,
and transform them into $P'$ and $Q'$ such that the diagram in
Fig.~\ref{fig:pushouts} commutes. For more
\begin{wrapfigure}{l}{.4\textwidth}
\centering
\includegraphics[scale=0.35]{Figures/pushouts-2}
\caption{Commutative diagram for Def.~\ref{def:mergeable-type-2}.}
\label{fig:pushouts-2}
\end{wrapfigure}
sophisticated data types, such as list, this process is more involved,
and may require the programmer to perform non-trivial reasoning. We
now present an alternative definition of a mergeable type that is
stronger than Def.~\ref{def:mergeable-type}, but also serves as a
recipe to build non-trivial merge functions, like that of a list.

\paragraph{Notation} As mentioned earlier, we distinguish between the
named functions defined by a data type library (primitive morphisms),
and their compositions (morphisms). A sequence of primitive morphisms
is written $P^*$ or $Q^*$. The composition of a sequence of primitive
morphisms (written $\denot{P^*}$) is their right-to-left composition,
i.e., $\C{fold\_left}\;(\lambda f.\lambda g. f \circ g)\; id\; P^*$,
where $\C{fold\_left}$ has its usual OCaml type. The length of a
sequence $S$ is written $\length{S}$.

@@

The definition asserts the existence of two functions for a mergeable
type $t$. First, the function $W$, which (implicitly) computes the
edit distance between an object $A:t$ and its semantic successor
$B:t$.  The sequence of primitive morphisms it returns ($P^*$) is
called an edit sequence. Applying the edit sequence on $A$ produces
$B$ (i.e., $\denot{P^*}(A) = B$). Second, the definition asserts the
existence of a function $M$, which performs an operational
transformation of the primitive morphisms in an edit script. An
operational transformation of a primitive morphism $P$ w.r.t a
morphism $Q$ is another primitive morphism $P'$ such that for any
object $A:t$, $P'$ has the same \emph{effect} on $Q(A)$ as $P$ has on
$A$.

Def.~\ref{def:mergeable-type-2} is immediately applicable to the list
type, and lets us explain why it is a mergeable type.  \C{MList.edit\_seq}
and \C{MList.op\_transform}, respectively, are the evidences for the
existence of $W$ and $M$ functions.

### 2017-02-27 paper/meta.tex

To define the semantics of LCA, we formalize the intuitive notation of
the ancestor relationship between versions of a legal branching
history (i.e., a branching history generated by the rules in
Fig.~\ref{fig:opsem}):

\begin{definition} [\bfseries Ancestor]
Version $v_1$ is a ancestor of version $v_2$ under a history
$H$ (written $\under{H}{v_1 \preceq v_2}$) if and only if one of the
following is true:
\begin{itemize}
  \item There exists a branch $b$ in $H$ (i.e., $\exists(t\in
  dom(H)).\,H(t) = b$) in which $v_2$ immediately succeeds
  $v_1$,
  \item There exists a branch $b$ in $H$ that contains $(v_2, 
  \C{FORK}\; (v_1,f_1)::b_1)$, for some $f_1$ and $b_1$,
  \item There exists a branch $b$ in $H$ that contains
  $(v_2, \C{MERGE}\;(v_1,f_1)::b_1)$, for some $f_1$ and $b_1$,
  \item $v_1 = v_2$, or $v_1$ is transitively a ancestor of
  $v_2$, i.e., $\exists v.~ \under{H}{v_1 \preceq v} \conj
  \under{H}{v \preceq v}$ 
\end{itemize}
\end{definition}

The ancestor relation is therefore a partial order with a greatest
lower bound (the initial version).  Thus, for any two versions in a
legal history, there exists at least one common ancestor.  The ancestor
relationship among all common ancestors lets us define the notion of a
least common ancestor (LCA):

\begin{definition} [\bfseries Least Common Ancestor]
Version $v$ is said to be a common ancestor of versions $v_1$ and
$v_2$ under a history $H$ if and only if $\under{H}{v \preceq v_1}$
and $\under{H}{v \preceq v_2}$. It is said to be the least common
ancestor (LCA) of $v_1$ and $v_2$, iff there does not exist a $v'$
such that $\under{H}{v' \preceq v_1}$ and $\under{H}{v' \preceq v_2}$
and $\under{H}{v \preceq v'}$.
\end{definition}

When merging two concurrent versions $v_1$ and $v_2$, the common
ancestor argument for \C{merge} must be the LCA of $v_1$ and $v_2$,
without which \C{merge} may yield unexpected results. This is
demonstrated for the grow-only counter in
Fig.~\ref{fig:merge-needs-lca}, where an incorrect count is obtained if a
common ancestor that is not an LCA is used to merge 4 and 7. While in
this example there is a unique LCA for 4 and 7, in general this may
\begin{wrapfigure}{l}{.4\textwidth}
\centering
\includegraphics[scale=0.6]{Figures/merge-needs-lca}
\caption{This example of a grow-only counter illustrate why \C{merge}
needs a least common ancestor, and not just a common ancestor. Both 0
and 2 are common ancestors of 4 and 7, while 2 is their least common
ancestor (since $0 \preceq 2$). The result (v) of merging 4 and 7 is
11 (incorrect) if 0 is used as the common ancestor for merge, and 9
(correct, because 2+2+3+2 = 9) if 2 is used. }
\label{fig:merge-needs-lca}
\end{wrapfigure}
not be the case. With unrestrained branching and merging, there is no
bound on the number of LCAs a pair of versions can have.  For example,
in Fig.~\ref{fig:criss-cross-lcas}, the merge of 0 with 3 is preceded
by two ``criss-cross'' merges between their respective branches
resulting in there being two LCAs (5 and 4) for 0 and 3. Multiple LCAs
can occur even without criss-cross merges, as demonstrated by
Fig.~\ref{fig:external-lcas}. Concurrent versions with multiple LCAs
do not lend themselves to three-way merging. This problem also arises
in the context of source control systems, which employ \emph{ad hoc}
mechanisms to pave the way for three-way merging.
GitHub~\cite{github}, for instance, recursively merges LCAs to compute
a virtual ancestor, which then serves as the LCA for merging
concurrent versions. This method is demonstrated in the branching
structure for a mergeable, replicated counter as shown in
Fig.~\ref{fig:criss-cross-lcas}, where LCAs 5 and 4 of 0 and 3 are
merged (with their LCA being 10) to generate -1 as the virtual LCA to
merge 5 and 4. A major downside with this approach is that it makes no
guarantees on the relationship between the virtual ancestor and its
concurrent versions; the former may not even be a legal ancestor of
the latter as per the semantics of the data type. For instance,
suppose the integer type in Fig.~\ref{fig:criss-cross-lcas} represents
a bank account balance, which is expected to disallows any activity on
the account if the balance is ever known to be less than zero.  From
the perspective of the library designer and its clients, there is no
meaningful scenario in which versions 3 and 0 can emerge from -1,
since the only transition allowed by the semantics from -1 is to
itself.  Clearly, \emph{ad hoc} mechanisms like this are error-prone
and difficult to apply in general.

\begin{figure}[!t]
\centering
\subcaptionbox[] {\small
  In this example, 1 and 3 have two LCAs (3 and 4) a result of
  previous merges. The dotted circle denotes a virtual ancestor
  obtained by merging the two LCAs.
  \label{fig:criss-cross-lcas}
} [0.47\columnwidth] {
  \includegraphics[scale=0.55]{Figures/2-LCAs}
}
\hfill
\subcaptionbox[] {\small
  In this example, versions $v_{13}$ and $v_{44}$ have two LCAs
  ($v_{22}$ and $v_{32}$)  despite there not being any previous merges
  between their respective branches.
  \label{fig:external-lcas}
} [0.47\columnwidth] {
  \includegraphics[scale=0.7]{Figures/2-external-LCAs}
}
\caption{Examples where merging versions have more than one LCA}
\label{fig:many-lcas}
\end{figure}

Fortunately, unlike source control systems where branching structure
is entirely dictated by the user, \name abstracts away branching
structure from the programmer, and hence retains the ability to
manifest it in a way that it deems fit. In particular, \name solves
the problem of multiple LCAs either by constraining the branching
stucture to prevent their existence, or by merging branches to ensure
that they have a new unique LCA.  Preention is done for
criss-cross merges by the operational semantics, which only allows
merging versions that are the latest on their respective branches. For
instance, suppose $v_{11}$ and $v_{21}$ are the latest versions on branches
$b_1$ and $b_2$, respectively. Merging $b_2$ into $b_1$ entails
merging $v_{21}$ into $v_{11}$ to generate version $v_{12}$ on $b_1$.
Now, merging $b_1$ into $b_2$ involves merging $v_{12}$, the latest
version on $b_1$, into $v_{21}$, but not $v_{11}$ into $v_{21}$, thus
preempting a criss-cross branching structure.
% In other words, a criss-cross branching structure is prevented due
% to the order among merges between conflicting branches introduced as
% a result of \rulelabel{E-Pull-Wait} being an atomic step.

Next, for branching structures where a pair of branches (i.e., their
respective latest versions) have multiple externally-located LCAs, we
suitably extend the branching structure to generate a new single LCA.
 For instance, consider
the example in Fig.~\ref{fig:external-lcas}, where (latest versions
on) branches $b_1$ and $b_4$ have two externally-located LCAs
($v_{22}$ and $v_{32}$), hence cannot be merged. Mergeability can
however be reclaimed by extending the branching structure as shown in
Fig.~\ref{fig:legal-extension-1}. The extension is done by first
merging $b_3$ into $b_2$ to create version $v_{23}$ on $b_2$, then
merging $b_2$ into $b_1$ to create version $v_{14}$ on $b_1$, and
finally merging $b_2$ into $b_4$ to create version $v_{45}$ on
$b_4$. Observe that $b_1$ and $b_4$ now have a single LCA ($v_{23}$)
hence are now mergeable. Mergeability could also have been reclaimed
by adopting an alternative extension to the branching structure as
shown in Fig.~\ref{fig:legal-extension-2}.

\begin{figure}[!t]
\centering
\subcaptionbox[] {\small
  $b_3$ is merged into $b_2$ to create $v_{23}$ here.
  \label{fig:legal-extension-1}
} [0.47\columnwidth] {
  \includegraphics[scale=0.8]{Figures/legal-extension-1}
}
\hfill
\subcaptionbox[] {\small
  $b_2$ is merged into $b_3$ to create $v_{33}$ here.
  \label{fig:legal-extension-2}
} [0.47\columnwidth] {
  \includegraphics[scale=0.8]{Figures/legal-extension-2}
}
\caption{Legal extensions of the branching structure in
  Fig.~\ref{fig:external-lcas} that reclaim the mergeability of $b_1$
  and $b_4$}
\label{fig:legal-extensions}
\end{figure}

Note that the extension strategy described above is possible only
because operational semantics imposes no obligation to merge any two
concurrent versions. The high-level purpose of merging branches in
\name is to propagate the data across the branches, which can be done
by merging any two concurrent versions that contain the data; the
semantics lets us pick these versions. While mergeability of branches
with multiple LCAs can be reclaimed by actively extending the
branching structure as described above, the semantics is
\emph{self-correcting} in the sense that a fair execution reclaims the
mergeability of any two branches without any intervention. This is
possible because the system preempts criss-cross merges, thus letting
the semantics make progress from any state. The property is formalized
by the following theorem\footnote{The more precise statement of the
theorem, and its proof are included in the appendix}:

\begin{theorem} [\bfseries Semantics Self-Correcting]
??
\end{theorem}

The \rulelabel{E-Pull-Wait} uses the function \C{lca} to compute the
LCA of (the latest versions on) a pair of branches. The definition of
the function is standard, hence not discussed.

>>>>

\subsection{Properties of the LCA}

When merging two concurrent versions $v_1$ and $v_2$, the common
ancestor argument for \C{merge} must be the LCA of $v_1$ and $v_2$,
without which \C{merge} may yield unexpected results. This is
demonstrated for the grow-only counter in
Fig.~\ref{fig:merge-needs-lca}, where an incorrect count is obtained if a
common ancestor that is not an LCA is used to merge 4 and 7. While in
this example there is a unique LCA for 4 and 7, in general this may
\begin{wrapfigure}{l}{.4\textwidth}
\centering
\includegraphics[scale=0.6]{Figures/merge-needs-lca}
\caption{This example of a grow-only counter illustrate why \C{merge}
needs a least common ancestor, and not just a common ancestor. Both 0
and 2 are common ancestors of 4 and 7, while 2 is their least common
ancestor (since $0 \preceq 2$). The result (v) of merging 4 and 7 is
11 (incorrect) if 0 is used as the common ancestor for merge, and 9
(correct, because 2+2+3+2 = 9) if 2 is used. }
\label{fig:merge-needs-lca}
\end{wrapfigure}
not be the case. With unrestrained branching and merging, there is no
bound on the number of LCAs a pair of versions can have.  For example,
in Fig.~\ref{fig:criss-cross-lcas}, the merge of 0 with 3 is preceded
by two ``criss-cross'' merges between their respective branches
resulting in there being two LCAs (5 and 4) for 0 and 3. 
% Multiple LCAs can occur even without criss-cross merges, as
% demonstrated by Fig.~\ref{fig:external-lcas}. 
Concurrent versions with multiple LCAs do not lend themselves to
three-way merging. If such versions are latest on their respective
branches, they render the branches unmergeable (since  \C{lca} is no
longer a function). This problem also arises in the context of source
control systems, which employ \emph{ad hoc} mechanisms to pave the way
for three-way merging.  GitHub~\cite{github}, for instance,
recursively merges LCAs to compute a virtual ancestor, which then
serves as the LCA for merging concurrent versions. This method is
demonstrated in the branching structure for a mergeable, replicated
counter as shown in Fig.~\ref{fig:criss-cross-lcas}, where LCAs 5 and
4 of 0 and 3 are merged (with their LCA being 10) to generate -1 as
the virtual LCA to merge 5 and 4. A major downside with this approach
is that it makes no guarantees on the relationship between the virtual
ancestor and its concurrent versions; the former may not even be a
legal ancestor of the latter as per the semantics of the data type.
For instance, suppose the integer type in
Fig.~\ref{fig:criss-cross-lcas} represents a bank account balance,
which is expected to disallows any activity on the account if the
balance is ever known to be less than zero.  From the perspective of
the library designer and its clients, there is no meaningful scenario
in which versions 3 and 0 can emerge from -1, since the only
transition allowed by the semantics from -1 is to itself.  Clearly,
\emph{ad hoc} mechanisms like this are error-prone and difficult to
apply in general.

\begin{figure}[!t]
\centering
\subcaptionbox[] {\small
  In this example, 1 and 3 have two LCAs (3 and 4) a result of
  previous merges. The dotted circle denotes a virtual ancestor
  obtained by merging the two LCAs.
  \label{fig:criss-cross-lcas}
} [0.47\columnwidth] {
  \includegraphics[scale=0.55]{Figures/2-LCAs}
}
% \hfill
% \subcaptionbox[] {\small
%   In this example, versions $v_{13}$ and $v_{44}$ have two LCAs
%   ($v_{22}$ and $v_{32}$)  despite there not being any previous merges
%   between their respective branches.
%   \label{fig:external-lcas}
% } [0.47\columnwidth] {
%   \includegraphics[scale=0.7]{Figures/2-external-LCAs}
% }
\caption{Examples where merging versions have more than one LCA}
\label{fig:many-lcas}
\end{figure}

Fortunately, unlike source control systems where branching structure
is entirely dictated by the user, \name abstracts away branching
structure from the programmer, and hence retains the ability to
manifest it in a way that it deems fit. In particular, \name solves
the problem of criss-cross merges by constraining the branching
stucture to prevent their existence. As captured in the operational
semantics, this is done by restricting the merge operation only to the
tips (latest versions) of the branches. For instance, suppose $v_{11}$
and $v_{21}$ are the latest versions on branches $b_1$ and $b_2$,
respectively. Merging $b_2$ into $b_1$ entails merging $v_{21}$ into
$v_{11}$ to generate version $v_{12}$ on $b_1$.  Now, merging $b_1$
into $b_2$ involves merging $v_{12}$, the latest version on $b_1$,
into $v_{21}$, but not $v_{11}$ into $v_{21}$, thus preempting a
criss-cross branching structure.
% In other words, a criss-cross branching structure is prevented due
% to the order among merges between conflicting branches introduced as
% a result of \rulelabel{E-Pull-Wait} being an atomic step.

% Next, for branching structures where a pair of branches (i.e., their
% respective latest versions) have multiple externally-located LCAs, we
% suitably extend the branching structure to generate a new single LCA.
%  For instance, consider
% the example in Fig.~\ref{fig:external-lcas}, where (latest versions
% on) branches $b_1$ and $b_4$ have two externally-located LCAs
% ($v_{22}$ and $v_{32}$), hence cannot be merged. Mergeability can
% however be reclaimed by extending the branching structure as shown in
% Fig.~\ref{fig:legal-extension-1}. The extension is done by first
% merging $b_3$ into $b_2$ to create version $v_{23}$ on $b_2$, then
% merging $b_2$ into $b_1$ to create version $v_{14}$ on $b_1$, and
% finally merging $b_2$ into $b_4$ to create version $v_{45}$ on
% $b_4$. Observe that $b_1$ and $b_4$ now have a single LCA ($v_{23}$)
% hence are now mergeable. Mergeability could also have been reclaimed
% by adopting an alternative extension to the branching structure as
% shown in Fig.~\ref{fig:legal-extension-2}.

\begin{figure}[!t]
\centering
\subcaptionbox[] {\small
  In this example, versions $v_{13}$ and $v_{44}$ have two LCAs
  ($v_{22}$ and $v_{32}$) rendering their respective branches
  unmergeable.
  \label{fig:external-lcas}
} [0.47\columnwidth] {
  \includegraphics[scale=0.7]{Figures/2-external-LCAs}
}
\hfill
\subcaptionbox[] {\small
  However, execution can still progress to the depicted state, where
  $b_1$ and $b_4$ are again mergeable.
  \label{fig:legal-extension-1}
} [0.47\columnwidth] {
  \includegraphics[scale=0.8]{Figures/legal-extension-1}
}
% \hfill
% \subcaptionbox[] {\small
%   $b_2$ is merged into $b_3$ to create $v_{33}$ here.
%   \label{fig:legal-extension-2}
% } [0.47\columnwidth] {
%   \includegraphics[scale=0.8]{Figures/legal-extension-2}
% }
\caption{Multiple LCAs without criss-cross merges do not threaten
the progress guarantee.}
\label{fig:legal-extensions}
\end{figure}

There can be other scenarios that lead to multiple LCAs, such as the
one shown in Fig.~\ref{fig:external-lcas}. However, preemption of
criss-cross merges is sufficient to make progress despite multiple
LCAs. For instance, the branching history in
Fig.~\ref{fig:external-lcas}, where $b_1$ and $b_4$ cannot merge due
to multiple LCAs, can still make progress to the history shown in
~\ref{fig:legal-extension-1}, where $b_1$ and $b_4$ are now mergeable.
We formalize the progress guarantee in the following theorem, whose
proof is included in the appendix.

\begin{theorem} [\bfseries Progress]
In a legal branching history $H$ produced by the operational
semantics, there exist at least two branches that have a unique LCA,
hence mergeable.
\end{theorem}

### 2017-02-27 paper/model.tex

\section{Programming Model}

@@

{\bf Shopping List}. A non-trivial demonstration of composability of
mergeable types is a shopping list application. The application allows
its users to collaboratively build a grocery shopping list (listing
items by decreasing quantity), by adding new items with a specified
quantity, deleting existing items, and increasing or decreasing their
quantity. An item (\C{Item.t}) is represented as a tuple of its name
(a string) and the quantity (A \C{Qty.t}). The merge function for
\begin{wrapfigure}{l}{.5\textwidth}
  \centering
  \includegraphics[scale=0.3]{Figures/shopping-list}
  \caption{Merging concurrent shopping lists}
  \label{fig:shopping-list}
\end{wrapfigure}
\C{Item.t} merges the quantities of items (using \C{Qty.merge}) having
the same names. Items with different names are considered distinct. A
shopping list is a mergeable list of items (i.e.,
\C{MList(Item)}). \C{MList(Item).merge} automatically lifts the merge
semantics for items to merge semantics for lists of items.
Fig.~\ref{fig:shopping-list} illustrates its behavior.  Here, two
users, Alice and Bob, concurrently edit a shopping list whose initial
contents are \C{milk} and \C{eggs}. While both simultaneously update the
quantity of eggs, Bob also removes milk and inserts candy. The merged
shopping list contains eggs and candy, where the quantity of eggs is
obtained by merging Alice's and Bob's quantity as per the definition
of \C{Qty.merge}.

>>>>

\section{Programming Model}
\label{sec:model}

@@

{\bf Shopping List}. A non-trivial demonstration of composability of
mergeable types is a shopping list application. The application allows
its users to collaboratively build a grocery shopping list (listing
items in the decreasing order of their priority), by adding new items
with a specified quantity, deleting existing items, and increasing or
decreasing their quantity. An item (\C{Item.t}) is represented as a
tuple of its name (a string) and the quantity (A \C{Qty.t}). The merge
function for
\begin{wrapfigure}{l}{.5\textwidth}
  \centering
  \includegraphics[scale=0.3]{Figures/shopping-list}
  \caption{Merging concurrent shopping lists}
  \label{fig:shopping-list}
\end{wrapfigure}
\C{Item.t} merges the quantities of items (using \C{Qty.merge}) having
the same names. Items with different names are considered distinct. A
shopping list is a mergeable list of items (i.e.,
\C{MList(Item)}). \C{MList(Item).merge} automatically lifts the merge
semantics for items to merge semantics for lists of items.
Fig.~\ref{fig:shopping-list} illustrates its behavior.  Here, two
users, Alice and Bob, concurrently edit a shopping list whose initial
contents are \C{milk} and \C{eggs}. While both simultaneously update the
quantity of eggs, Bob also removes milk and inserts candy. The merged
shopping list contains eggs and candy, where the quantity of eggs is
obtained by merging Alice's and Bob's quantity as per the definition
of \C{Qty.merge}.

### 2017-02-27 paper/semantics.tex

\section{Operational Semantics}

@@

The semantics non-deterministically chooses \rulelabel{E-Pull} or
\rulelabel{E-Pull-Wait} rules to reduce a \C{pull} expression. The
\rulelabel{E-Pull} rule reduces \C{pull} to \C{()}, and returns the
latest version on the current branch. The \rulelabel{E-Pull-Wait} rule
can be thought of as a stutter step; it doesn't reduce \C{pull}, but
updates the branching structure by merging (the latest version of) a
concurrent branch ($H(t')$) into (the latest version of) the current
branch ($H(t)$), and extending the current branch with the merged
version ($v_m$). The new version is tagged with a \C{MERGE} tag that,
like a \C{FORK} tag, records its origin. The rule assumes a function
\C{lca} that computes the \emph{least common ancestor} (LCA) for the
latest versions on the given pair of branches. The semantics of LCA is
discussed below. The \rulelabel{E-Pull-Wait} and \rulelabel{E-Pull}
rules thus let a thread synchronize with other distributed threads
manipulating versions of replicas in multiple steps before returning
the result of the \C{pull}. Since \C{sync} is a composition of
\C{push} and \C{pull}, its behavior can be explained thus: \C{sync}
pushes the given value onto the current (local) branch, merges a
(possibly empty) subset of concurrent branches into the local branch,
and returns the result.  (The operation is not guaranteed to
synchronize with all concurrent branches because not all such branches
may be available.)

@@

\input{system-model}

>>>>

\section{Formalization}

\subsection{Operational Semantics}

@@

The semantics non-deterministically chooses \rulelabel{E-Pull} or
\rulelabel{E-Pull-Wait} rules to reduce a \C{pull} expression. The
\rulelabel{E-Pull} rule reduces \C{pull} to \C{()}, and returns the
latest version on the current branch. The \rulelabel{E-Pull-Wait} rule
can be thought of as a stutter step; it doesn't reduce \C{pull}, but
updates the branching structure by merging (the latest version of) a
concurrent branch ($H(t')$) into (the latest version of) the current
branch ($H(t)$), and extending the current branch with the merged
version ($v_m$). The new version is tagged with a \C{MERGE} tag that,
like a \C{FORK} tag, records its origin. The rule assumes a function
\C{lca} that computes the \emph{least common ancestor} (LCA) for the
latest versions on the given pair of branches. The formal definition
of LCA is given below. The \rulelabel{E-Pull-Wait} and
\rulelabel{E-Pull} rules thus let a thread synchronize with other
distributed threads manipulating versions of replicas in multiple
steps before returning the result of the \C{pull}. Since \C{sync} is a
composition of \C{push} and \C{pull}, its behavior can be explained
thus: \C{sync} pushes the given value onto the current (local) branch,
merges a (possibly empty) subset of concurrent branches into the local
branch, and returns the result.  (The operation is not guaranteed to
synchronize with all concurrent branches because not all such branches
may be available.)

To define LCA, we formalize the intuitive notation of the ancestor
relationship between versions of a legal branching history (i.e., a
branching history generated by the rules in Fig.~\ref{fig:opsem}):

\begin{definition} [\bfseries Ancestor]
Version $v_1$ is a ancestor of version $v_2$ under a history
$H$ (written $\under{H}{v_1 \preceq v_2}$) if and only if one of the
following is true:
\begin{itemize}
  \item There exists a branch $b$ in $H$ (i.e., $\exists(t\in
  dom(H)).\,H(t) = b$) in which $v_2$ immediately succeeds
  $v_1$,
  \item There exists a branch $b$ in $H$ that contains $(v_2, 
  \C{FORK}\; (v_1,f_1)::b_1)$, for some $f_1$ and $b_1$,
  \item There exists a branch $b$ in $H$ that contains
  $(v_2, \C{MERGE}\;(v_1,f_1)::b_1)$, for some $f_1$ and $b_1$,
  \item $v_1 = v_2$, or $v_1$ is transitively a ancestor of
  $v_2$, i.e., $\exists v.~ \under{H}{v_1 \preceq v} \conj
  \under{H}{v \preceq v}$ 
\end{itemize}
\end{definition}

The ancestor relation is therefore a partial order with a greatest
lower bound (the initial version).  Thus, for any two versions in a
legal history, there exists at least one common ancestor.  The ancestor
relationship among all common ancestors lets us define the notion of a
least common ancestor (LCA):

\begin{definition} [\bfseries Least Common Ancestor]
Version $v$ is said to be a common ancestor of versions $v_1$ and
$v_2$ under a history $H$ if and only if $\under{H}{v \preceq v_1}$
and $\under{H}{v \preceq v_2}$. It is said to be the least common
ancestor (LCA) of $v_1$ and $v_2$, iff there does not exist a $v'$
such that $\under{H}{v' \preceq v_1}$ and $\under{H}{v' \preceq v_2}$
and $\under{H}{v \preceq v'}$.
\end{definition}

The function \C{lca} used in the premise of \rulelabel{E-Pull-Wait}
computes the LCA of two branches. The function is in fact a partial
function because the definition of LCA doesn't guarantee its
uniqueness for a pair of branches. Such nuances of LCA, and the
guarantees provided by the system notwithstanding these nuances, are
discussed in the subsection~\ref{sec:meta}.

\input{mergeable-types}

@@
